---
title: "Validation of temperature-related mortality curves for Italian cities"
subtitle: "Masselot et al. (2023) exposure-response functions, Istat daily deaths and Open-Meteo forecasts"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
    fig-width: 9
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(targets)
library(data.table)
library(ggplot2)
theme_set(theme_minimal(base_size = 12))
cfg      <- tar_read(config)
periods  <- tar_read(periods)
data_end <- tar_read(data_end)
xw       <- tar_read(crosswalk)
fmt <- function(x, d = 2) formatC(x, format = "f", digits = d, big.mark = ",")
pct <- function(x, d = 0) paste0(fmt(100 * x, d), "%")
kbl <- function(x, ...) knitr::kable(x, ...)
```

# Purpose and design

The published exposure-response functions (ERFs) of Masselot et al. (2023,
*Lancet Planetary Health*) are the standard source of temperature-mortality
risk for European cities, and the natural basis for a heat-mortality
forecasting system. This report asks whether they still describe mortality in
Italian cities after the period they were estimated on, how much of any
mismatch can be repaired by re-estimating them on recent Italian data, and
how far ahead a temperature forecast can drive them.

Masselot's curves come from a three-stage model. In 232 cities with observed
daily mortality (for Italy, the MCC data covering 2001-2010), a quasi-Poisson
distributed-lag non-linear model was fitted; the city curves were then pooled
in a meta-regression on city characteristics, which was used to predict
curves for all 854 cities. The paper does not include an out-of-sample test
over later years. We therefore validate the curves in the way they were
built, plus a check aimed at the forecasting use:

1. **Temperature forecast skill** — archived Open-Meteo forecasts
   (`r cfg$forecast_model`, leads `r min(cfg$forecast_leads)`-`r max(cfg$forecast_leads)`
   days) against ERA5-Land, to decide the usable horizon.
2. **Curve comparison (test A)** — Masselot's first-stage model refitted,
   with identical specification, on Istat deaths and ERA5-Land for
   `r periods$test$label`, and compared with the published curves
   `r if (!is.null(periods$repro)) paste0("(and for ", periods$repro$label, ", a period also independent of the Italian MCC data)") else ""`.
3. **Observed vs predicted heat deaths (test B)** — for heatwave episodes
   and whole summers, the heat deaths predicted by the curves are compared
   with the deaths observed above a counterfactual baseline of deaths
   expected without heat, for the published curves and for curves
   re-estimated on Italian data of an earlier period.

Mortality data end on **`r format(data_end, "%d %B %Y")`**.

# Data and alignment

## Do we reproduce the paper's own numbers?

Everything below rests on our reconstruction of the published curves, so it
is checked first: our code is run on Masselot's own exposure series
(1990-2019) and his own annual deaths, and compared with the numbers
published in `results.zip`. Minimum mortality temperatures come out
identical and annual heat-attributable deaths are reproduced exactly, so the
curves, the MMT search and the attribution formula are applied as in the
paper and any mismatch below belongs to the curves rather than to us. The
relative risk at the 99th percentile differs because the paper reports its
curve on a spline ending at the 0.1st and 99.9th percentiles while
attribution uses one ending at the daily minimum and maximum; comparisons
here always use the same basis on both sides.

```{r recon}
rc <- tar_read(recon_check)
kbl(rc[, .(`City x age` = .N,
           `Median MMT difference (°C)` = fmt(median(mmt_diff)),
           `Max |MMT difference|` = fmt(max(abs(mmt_diff))),
           `Median ratio of excess risk at P99` = fmt(median(rr_ratio), 3),
           `Median ratio of annual heat deaths` = fmt(median(heat_ratio), 3))],
    caption = "Our reconstruction vs the published results (1 = identical)")
```

```{r recon-plot, fig.height=4}
#| fig-cap: "Each point is a city and age group over 1990-2019."

ggplot(rc, aes(heat_pub, heat_ours, colour = agegroup)) + geom_abline(colour = "grey50") +
  geom_point(alpha = .7) + scale_x_log10() + scale_y_log10() +
  labs(x = "Annual heat deaths published by Masselot et al.", y = "Recomputed here",
       colour = "Age", title = "Reconstruction check, Italian cities, 1990-2019")
```

The `r nrow(tar_read(recon_check))` city-age combinations sit on the diagonal
across three orders of magnitude, from cities with a handful of heat deaths a
year to the largest. The pipeline therefore is the paper's pipeline.

## City definition

Masselot's cities are Urban Audit cities, coded with Eurostat's 2020/2021
numbering, in which the *greater city* is the default unit wherever the
urban centre extends well beyond the central municipality (e.g. Milan,
Naples): from the 2021 release greater cities are merged into the city
level, so a single "city" polygon can cover dozens of comuni. Urban Audit
codes were also renumbered between releases, so matching polygons by code
can silently select another city. Cities are therefore
matched geographically: each city's Urban Audit point (from Masselot's
metadata) is located in the GISCO `r cfg$urau_year` polygons of every level
(`r paste(cfg$urau_levels, collapse = ", ")`); each polygon containing the
point is a candidate, made of the comuni with at least
`r pct(cfg$lau_share_min)` of their area inside it. The candidate whose
population is closest to Masselot's is retained. The choice is then checked
against the Istat deaths aged 20+ in 2011-2019 and the deaths implied by
Masselot's population structure and death rates: cities failing this check
are excluded, since their mortality would not be the population the curve
describes.

```{r crosswalk-summary}
ck <- xw$checks
kbl(ck[, .N, by = .(status, analysed)][order(-N)],
    col.names = c("Status", "Analysed", "Cities"))
kbl(ck[, .N, by = .(level)][order(-N)], col.names = c("Definition retained", "Cities"))
```

```{r crosswalk-greater}
gc <- ck[level != "CITIES" & !is.na(level),
         .(City = name, Polygon = poly_name, Level = level, Comuni = n_comuni,
           `Pop. ratio` = fmt(pop_ratio), `Deaths ratio` = fmt(deaths_ratio), Status = status)]
if (nrow(gc)) kbl(gc, caption = "Cities defined as greater cities (or FUA) rather than the single comune")
```

```{r crosswalk-plot, fig.height=4.5}
ggplot(ck[!is.na(deaths_ratio)], aes(pop_ratio, deaths_ratio, colour = status)) +
  annotate("rect", xmin = cfg$pop_ratio_ok[1], xmax = cfg$pop_ratio_ok[2],
           ymin = cfg$deaths_ratio_ok[1], ymax = cfg$deaths_ratio_ok[2],
           alpha = .08, fill = "darkgreen") +
  geom_point(aes(size = pop), alpha = .7) +
  geom_text(data = ck[!is.na(deaths_ratio) & status != "ok"], aes(label = name),
            size = 3, vjust = -1, show.legend = FALSE) +
  scale_x_log10() + scale_y_log10() + scale_size_area(guide = "none") +
  labs(x = "Population of the retained comuni / Masselot population",
       y = "Istat deaths 20+ / deaths implied by metadata",
       colour = NULL, title = "Alignment of city definitions (green = accepted range)")
```

```{r crosswalk-flagged}
#| tbl-cap-location: top

fl <- ck[status != "ok", .(City = name, Polygon = poly_name, Level = level, Status = status,
                           Candidates = n_candidates, `Pop. ratio` = fmt(pop_ratio),
                           `Deaths ratio` = fmt(deaths_ratio), Analysed = analysed)]
if (nrow(fl)) kbl(fl, caption = "Cities to inspect (a level or comuni can be forced in data/raw/crosswalk_overrides.csv)") else cat("All cities passed the checks.")
```

The population ratios now cluster between 0.9 and 1.1: the comuni we sum are
the population Masselot's curves describe. Five cities keep a deaths ratio
outside the accepted band while their population matches (Padova, Treviso,
Gela, Cerignola, Altamura); since the population is right, this reflects the
regional death rates in his metadata rather than a wrong city, and they stay
in the analysis with the flag. Anzio is retained with a population flag.
Cagliari and Sassari are dropped: comune codes in their definition are absent
from the Istat file, most likely because of code changes in Sardinia, and a
city with missing comuni would carry too few deaths.

## Age groups

Istat 5-year classes are merged into Masselot's groups (20-44, 45-64, 65-74,
75-84, 85+); deaths below age 20 are excluded, as in the paper.

```{r age-check}
ac <- tar_read(istat)$age_check
kbl(ac[, .(deaths = sum(deaths)), by = .(agegroup = fifelse(is.na(agegroup), "<20 (excluded)", agegroup))][
  , share := pct(deaths / sum(deaths), 1)], col.names = c("Age group", "Deaths (all years)", "Share"))
```

## Exposure

ERA5-Land daily mean temperature comes from Open-Meteo,
`r if (identical(cfg$era5_mode, "polygon")) sprintf("averaged over up to %d ERA5-Land cells inside each city polygon, as Masselot did", cfg$era5_max_points) else "taken at each city's point"`.
The two series are compared over `r format(cfg$overlap_start, "%Y")`-`r format(cfg$overlap_end, "%Y")`.
`r if (isTRUE(cfg$bias_correct)) "The city x month mean difference is removed from our series before any analysis, so temperatures sit on the scale of the distribution the ERF knots come from." else "No correction is applied."`

```{r exposure-check, fig.height=4}
ex <- tar_read(exposure_check)$summary
ex <- merge(ex, tar_read(cities)[, .(URAU_CODE, name)], by = "URAU_CODE")
kbl(ex[, .(Cities = .N, `Median bias (°C)` = fmt(median(bias)),
           `Max |bias| (°C)` = fmt(max(abs(bias))), `Median MAE (°C)` = fmt(median(mae)),
           `Median r` = fmt(median(r), 3))])
ggplot(ex, aes(bias, reorder(name, bias))) + geom_col(fill = "grey50") +
  labs(x = "Mean difference, our point series - Masselot polygon series (°C)", y = NULL) +
  theme(axis.text.y = element_text(size = 6))
```

```{r beyond-range}
#| include: true

td <- tar_read(erf)$tdist
pcs <- grep("%$", names(td), value = TRUE)
td[, tmax := do.call(pmax, .SD), .SDcols = pcs]
et <- merge(tar_read(exposure_test)[date >= periods$test$start], td[, .(URAU_CODE, tmax)], by = "URAU_CODE")
br <- et[, .(days = sum(tmean > tmax)), by = URAU_CODE][days > 0]
br <- merge(br, tar_read(cities)[, .(URAU_CODE, name)], by = "URAU_CODE")[order(-days)]
```

Averaging the cells inside each polygon brings the median difference to
`r fmt(ex[, median(bias)])` °C and the median daily error to
`r fmt(ex[, median(mae)])` °C, with a correlation of 1.000: our series and
Masselot's are effectively the same variable. What remains is topographic.
The largest positive differences are Trento, Cosenza and Bolzano, whose
polygons run up the surrounding slopes, so the city average is colder than
the built-up area; the largest negative ones, Lecco and Caserta, sit beside a
lake and a mountain range. The monthly mean difference removes what is left,
so the temperatures entering the curves are on the scale of the distribution
their knots come from.

In the test period, `r nrow(br)` cities had days warmer than the maximum of
their 1990-2019 distribution (`r sum(br$days)` city-days in total). On those
days both the published and the refitted curves are extrapolated beyond the
last spline knot, so relative risks there are uncertain; curves in the plots
below stop at the 1990-2019 maximum.

# Temperature forecast skill

```{r fc-setup}
fs <- tar_read(fc_skill)
```

Forecasts from the Open-Meteo Previous Runs API (value predicted N days
before the valid time; archive from `r format(cfg$forecast_start, "%B %Y")`)
are scored against ERA5-Land at the same point. The reference is persistence
(tomorrow = today): skill above 0 means the forecast beats it.

```{r fc-plot, fig.height=6}
sc <- melt(fs$scores, id.vars = c("subset", "lead"), measure.vars = c("mae", "rmse", "bias", "mae_persistence"))
sc[, variable := factor(variable, c("mae", "rmse", "bias", "mae_persistence"),
                        c("MAE", "RMSE", "Bias", "MAE of persistence"))]
ggplot(sc, aes(lead, value, colour = variable)) + geom_hline(yintercept = 0, colour = "grey70") +
  geom_line(linewidth = .8) + geom_point() + facet_wrap(~ subset) +
  scale_x_continuous(breaks = cfg$forecast_leads) +
  labs(x = "Lead time (days)", y = "°C", colour = NULL,
       title = "Error of daily mean temperature forecasts by lead time") +
  theme(legend.position = "bottom")
```

```{r fc-table}
kbl(dcast(fs$scores, lead ~ subset, value.var = "mae")[, lapply(.SD, fmt), by = lead],
    caption = "MAE (°C) by lead time")
kbl(fs$detection[, .(Lead = lead, Hits = hits, Misses = misses, `False alarms` = false_alarms,
                     POD = fmt(pod), FAR = fmt(far))],
    caption = sprintf("Detection of heatwave days (T >= P%s), warm season", cfg$hw_pct))
```

```{r fc-impact}
fi <- tar_read(fc_impact)
if (!is.null(fi)) {
  kbl(fi$summary[, .(Lead = lead, Episodes = episodes, `Total ratio forecast/ERA5` = fmt(ratio_total),
                     `MAE (deaths per episode)` = fmt(mae_deaths, 1),
                     `Median abs. relative error` = pct(median_abs_rel_err),
                     `Mean temp. error (°C)` = fmt(mean_temp_error))],
      caption = "Predicted excess deaths of each episode with forecast vs ERA5-Land temperatures")
}
```

```{r fc-numbers}
sc <- fs$scores[subset == "Warm season"][order(lead)]
hd <- fs$scores[subset == "Hot days (> P90)"][order(lead)]
dt <- fs$detection[order(lead)]
fi2 <- if (!is.null(fi)) fi$summary[order(lead)] else NULL
```

Forecast error grows slowly: the mean absolute error in the warm season is
`r fmt(sc$mae[1])` °C at one day and `r fmt(sc$mae[nrow(sc)])` °C at seven, and the
forecast beats persistence from day two onwards. On hot days it carries a warm
bias growing from `r fmt(hd$bias[1])` °C to `r fmt(hd$bias[nrow(hd)])` °C.
Heatwave days are detected reliably at every lead (probability of detection
`r fmt(min(dt$pod))`-`r fmt(max(dt$pod))`), at the cost of more false alarms as
the lead grows (`r fmt(min(dt$far))` to `r fmt(max(dt$far))`).

That warm bias matters more than its size suggests. Because the curve is
convex above the MMT, predicted heat deaths run
`r if (!is.null(fi2)) pct(fi2$ratio_total[1] - 1) else "well"` above the
ERA5-Land-based figure already at one day, rising to
`r if (!is.null(fi2)) pct(fi2$ratio_total[nrow(fi2)] - 1) else "more"` at seven. Most of
this is bias rather than noise, so it can be removed: `fc_debias` in the
configuration subtracts the city, month and lead mean error, which should be
switched on before the forecasts are used to produce numbers rather than
warnings. With it on, the usable horizon is set by the growth in error, not
by the bias.

**Reading the horizon.** Because the curve is convex above the MMT, random
forecast errors do not cancel out: they inflate predicted deaths on average,
so the total ratio drifts above 1 as the lead grows even without any
temperature bias. The error that matters is the one in predicted
deaths, since the curve is steep above the minimum-mortality temperature: the
last table translates temperature errors into episode deaths. A lead is
usable while its relative error stays well below the uncertainty of the curve
itself (the width of the predicted-excess interval in test B).

# Test A: refitting Masselot's first-stage model

In each city the model of Masselot's first stage is refitted exactly (script
`05_FirstStage.R` of the authors' repository): quasi-Poisson regression;
cross-basis with a quadratic B-spline of temperature (knots at the
10th/75th/90th percentiles and boundaries at the range of the city's
1990-2019 ERA5-Land distribution, i.e. the published basis) and a natural
spline of lags 0-`r cfg$lag_max` with `r cfg$lag_nknots` knots equally spaced
on the log scale; day-of-week indicators; natural spline of time with
`r cfg$df_time_per_year` df per year. The fit is reduced to the overall
cumulative curve, whose five coefficients are directly comparable with the
published ones. Series are all deaths 20+, and age groups merged with
Masselot's cumulative `r fmt(cfg$min_deaths, 0)`-death rule. The MMT is
searched between the `r cfg$mmp_range[1]`th and `r cfg$mmp_range[2]`th
percentiles, as in the paper.

The main comparison is the relative risk at the `r cfg$rr_pct`th percentile
versus the published MMT (a common reference for both curves). Differences
are pooled across cities with a random-effects meta-analysis.

```{r curve-fun}
show_curve_test <- function(ct, pooled, label) {
  r <- ct$results
  cat(sprintf("\n\n**%s.** %d cities fitted on all deaths 20+, %d not fitted (deaths below %s).\n\n",
              label, r[level == "all20" & status == "fitted", .N],
              r[level == "all20" & status != "fitted", .N], fmt(cfg$min_deaths, 0)))
  print(kbl(pooled[, .(Group = group, Cities = k,
                       `Ratio of RR99 (refit/published)` = fmt(ratio, 3),
                       `95% CI` = ifelse(is.na(lo), "", paste(fmt(lo, 3), "-", fmt(hi, 3))),
                       p = fmt(p, 3), `I²` = ifelse(is.na(i2), "", paste0(fmt(i2, 0), "%")))],
            caption = "Pooled difference in heat relative risk at P99 (1 = no difference)"))
}
forest_plot <- function(ct, pooled, label) {
  r <- ct$results[level == "all20" & status == "fitted"]
  r[, `:=`(lo = exp(d_lr99 - 1.96 * se_d_lr99), hi = exp(d_lr99 + 1.96 * se_d_lr99),
           src = fifelse(inmcc %in% TRUE, "Fitted on local data (MCC)", "Extrapolated"))]
  pool <- pooled[group == "All cities"]
  ggplot(r, aes(exp(d_lr99), reorder(name, d_lr99), colour = src)) +
    geom_vline(xintercept = 1, colour = "grey50") +
    annotate("rect", xmin = pool$lo, xmax = pool$hi, ymin = -Inf, ymax = Inf,
             alpha = .12, fill = "firebrick") +
    geom_vline(xintercept = pool$ratio, colour = "firebrick", linetype = 2) +
    geom_errorbarh(aes(xmin = lo, xmax = hi), height = 0) + geom_point() +
    scale_x_log10() +
    labs(x = "RR99 refitted / RR99 published (log scale)", y = NULL, colour = NULL,
         title = paste("City differences in heat risk,", label),
         subtitle = "Red band: pooled estimate and 95% CI") +
    theme(axis.text.y = element_text(size = 6), legend.position = "bottom")
}
```

## `r periods$test$label`

```{r test-a, results='asis'}
ct <- tar_read(curves_test); pt <- tar_read(pooled_test)
show_curve_test(ct, pt, periods$test$label)
```

```{r test-a-forest, fig.height=10}
forest_plot(ct, pt, periods$test$label)
```

Three and a half years are short for a model with 20 cross-basis parameters,
so city curves are drawn only over the temperatures observed in the period
and are stabilised as in Masselot's second stage: the refitted coefficients
are pooled in a multivariate meta-analysis and each city is represented by
its best linear unbiased prediction (BLUP), which borrows strength from the
other cities in proportion to its own imprecision.

```{r test-a-curves, fig.height=9}
bt <- tar_read(blup_test)
if (!is.null(bt)) {
  top <- ct$results[level == "all20" & status == "fitted"][order(-deaths)][1:min(12, .N), URAU_CODE]
  b <- merge(bt$bands[URAU_CODE %in% top], tar_read(cities)[, .(URAU_CODE, name)], by = "URAU_CODE")
  ggplot(b, aes(t, rr, colour = curve, fill = curve)) + geom_hline(yintercept = 1, colour = "grey60") +
    geom_ribbon(aes(ymin = lo, ymax = hi), alpha = .15, colour = NA) + geom_line(linewidth = .8) +
    facet_wrap(~ name, scales = "free_x") + coord_cartesian(ylim = c(.8, 2.5)) +
    scale_colour_manual(values = c(Published = "grey30", `Refitted (BLUP)` = "firebrick")) +
    scale_fill_manual(values = c(Published = "grey30", `Refitted (BLUP)` = "firebrick")) +
    labs(x = "Daily mean temperature (°C), range observed in the period",
         y = "Relative risk (vs own MMT)", colour = NULL, fill = NULL,
         title = "Published and refitted (BLUP) curves, deaths 20+, largest cities") +
    theme(legend.position = "bottom")
}
```

Heterogeneity between cities is small (I-squared of
`r fmt(pt[group == "All cities", i2], 0)`% in the test period), so this is a
common shift rather than a few outliers, and cities whose curves were fitted on
their own mortality show much the same gap as cities where the curve was
extrapolated from the meta-regression
(`r fmt(pt[grepl("MCC", group) & !grepl("Difference", group), ratio], 2)` against
`r fmt(pt[grepl("extrapolated", group) & !grepl("Difference", group), ratio], 2)`,
difference not significant). The spatial prediction is therefore not the
problem.

The disagreement is confined to the heat arm. Above the minimum the refitted
curve rises much less steeply than the published one in every large city, and
the gap widens with temperature. Below the minimum the two run close together
in most cities, while in several (Catania, Florence, Bari, Milan, Verona) the
refitted cold arm sits somewhat above the published one, so this is not a
curve that is flatter everywhere: it is a curve with a weaker heat effect.
The cold side is not what these tests are about, and it is confounded by
influenza, so no weight is put on it here.

```{r test-a-blup}
if (!is.null(bt)) kbl(bt$results[, .(Cities = .N, `Median RR99 published` = fmt(median(rr99_pub), 3),
                                    `Median RR99 refitted (BLUP)` = fmt(median(rr99_blup), 3),
                                    `Median ratio` = fmt(exp(median(d_lr99)), 3))],
                      caption = "Heat relative risk at P99 vs the published MMT, BLUP curves")
```

```{r test-a-mmt, fig.height=4.5}
r <- ct$results[status == "fitted"]
print(ggplot(r[level == "all20"], aes(af_pub, af_new)) + geom_abline(colour = "grey50") +
  geom_errorbar(aes(ymin = af_new - 1.96 * se_d_af, ymax = af_new + 1.96 * se_d_af),
                colour = "grey70", width = 0) + geom_point() +
  scale_x_continuous(labels = scales::percent) + scale_y_continuous(labels = scales::percent) +
  labs(x = "Heat-attributable fraction, published curve", y = "Refitted curve",
       title = sprintf("Heat-attributable fraction of deaths 20+, %s", periods$test$label)))
```

```{r test-a-age}
ra <- r[level == "age" & n_ages == 1 & !is.na(wald_p)]
if (nrow(ra)) kbl(ra[, .(Cities = .N, `Median RR99 ratio` = fmt(exp(median(d_lr99)), 3),
                         `Wald p < 0.05` = pct(mean(wald_p < .05))), by = series][order(series)],
                  caption = "Single age groups: equality of the five coefficients (Wald test)")
```

```{r test-a-repro, results='asis'}
if (isTRUE(cfg$run_reproduction)) {
cat("\n\n## ", periods$repro$label, "\n\n")
cat("This period uses Masselot's own ERA5-Land series, so the only difference ",
    "with the published model is the mortality data. For Italian MCC cities the ",
    "published curves used 2001-2010 data, so these years are independent too; ",
    "for extrapolated cities this is a direct test of the spatial prediction ",
    "over the period the curves describe.\n\n")
cr <- tar_read(curves_repro); pr <- tar_read(pooled_repro)
show_curve_test(cr, pr, periods$repro$label)
}
```

```{r test-a-repro-forest, fig.height=10}
if (isTRUE(cfg$run_reproduction)) forest_plot(cr, pr, periods$repro$label)
```

## Trend across time windows

The same first-stage model is refitted in successive windows. Windows ending
before 2020 use Masselot's own ERA5-Land series, so the exposure is identical
to his; later windows use our aligned series. A falling ratio means the heat
risk carried by the published curves is increasingly too steep for the data.

```{r periods}
cp <- tar_read(curves_periods)
kbl(cp$pooled[, .(Window = window, Cities = k, `Ratio of RR99 (refit/published)` = fmt(ratio, 3),
                  `95% CI` = paste(fmt(lo, 3), "-", fmt(hi, 3)), p = fmt(p, 3))],
    caption = "Pooled difference with the published curves, by time window")
if (!is.null(cp$trend))
  kbl(cp$trend[, .(`Change per decade` = fmt(change_per_decade, 3),
                   `95% CI` = paste(fmt(lo, 3), "-", fmt(hi, 3)), p = fmt(p, 3))],
      caption = "Meta-regression of the difference on the window midpoint")
```

```{r periods-plot, fig.height=4}
if (nrow(cp$pooled)) {
  ggplot(cp$pooled, aes(midyear, ratio)) +
    geom_hline(yintercept = 1, colour = "grey60") +
    geom_errorbar(aes(ymin = lo, ymax = hi), width = .3, colour = "grey50") +
    geom_point(size = 2.5, colour = "firebrick") +
    geom_text(aes(label = window), vjust = -1.4, size = 3) +
    labs(x = "Midpoint of the window", y = "Refitted / published heat risk at P99",
         title = "Is the gap widening over time?")
}
```

The ratio falls in every step, and the trend is significant. Two of the three
windows use Masselot's own temperatures, so the decline cannot be an artefact
of our exposure series.

# Test B: observed vs predicted heat deaths

**Deaths expected without heat.** For every city and age group a baseline
B is estimated: a quasi-Poisson model with day of week, a slow trend and
`r cfg$baseline_harmonics` annual harmonics, fitted only on *reference days*
— outside the influenza season (months
`r paste(month.abb[cfg$baseline_exclude_months], collapse = ", ")`) and not
within `r cfg$baseline_hot_lag` days after a day above the curve's MMT. Like
EuroMOMO baselines, it captures the normal seasonal pattern, including the
lower mortality of summer, without the heat itself; cool summer days anchor
the summer level. Reference days are cool, so part of their deaths is
cold-related: `r if (isTRUE(cfg$baseline_cold_offset)) "the published cold effect enters the model as an offset, so B is the number of deaths expected at the minimum-mortality temperature, the reference of the attributable-deaths formula" else "no correction is applied, so B also contains cold-related deaths"`. Age groups with fewer than `r cfg$baseline_min_daily`
deaths per day use the 20+ baseline times their share.

Over each window:

- **Observed heat deaths** X = observed deaths − B;
- **Predicted heat deaths** P = Σ B × (RR − 1) over the days at or above the
  MMT, with the relative risk centred on the MMT. Days below the MMT inside a
  window belong to the cold side of the curve and are counted separately.

If the curves are right, X scatters around P with slope 1. Windows are
heatwave episodes (spells of at least `r cfg$hw_min_days` days above the
city's P`r cfg$hw_pct`, plus `r cfg$tail_days` days), and whole summers
(June-September), which do not depend on any episode definition.
Uncertainty of both the baseline and the curves is propagated.

```{r test-b-setup}
cf <- tar_read(counterfactual); ev <- tar_read(cf_eval)
has_ep <- !is.null(ev)
```

```{r test-b-example, fig.height=4.5}
if (has_ep) {
  big <- ev$windows[type == "summer"][order(-O)][1]
  dd <- cf$daily[URAU_CODE == big$URAU_CODE & date >= big$start & date <= big$end,
                 .(O = sum(deaths), B = sum(B), E = sum(B * exp(logrr))), by = date]
  nm <- tar_read(cities)[URAU_CODE == big$URAU_CODE, name]
  dd <- melt(dd, id.vars = "date")
  dd[, variable := factor(variable, c("O", "B", "E"),
                          c("Observed", "Expected without heat (B)", "Predicted with published curve"))]
  print(ggplot(dd, aes(date, value, colour = variable)) +
    geom_line(data = dd[variable == "Observed"], alpha = .5) +
    geom_line(data = dd[variable != "Observed"], linewidth = .9) +
    scale_colour_manual(values = c("grey40", "steelblue", "firebrick")) +
    labs(x = NULL, y = "Deaths per day (20+)", colour = NULL,
         title = sprintf("%s, summer %s", nm, big$year)) +
    theme(legend.position = "bottom"))
}
```

```{r test-b-summary}
if (has_ep) {
  print(kbl(ev$calibration[, .(Windows = windows, N = n, `Observed heat deaths` = fmt(observed, 0),
                               `Predicted heat deaths` = fmt(predicted, 0),
                               `Observed / predicted` = fmt(ratio_total),
                               `Slope` = sprintf("%s (%s-%s)", fmt(slope), fmt(lo), fmt(hi)),
                               r = fmt(r), `Skill vs baseline` = pct(deviance_skill),
                               `95% PI coverage` = pct(coverage_95))],
            caption = "Calibration of the published curves (slope 1 = calibrated)"))
  print(kbl(ev$by_year[, .(Windows = type, Year = year, N = windows, Cities = cities,
                           Observed = fmt(O, 0), `Expected without heat` = fmt(B, 0),
                           `Observed heat deaths` = fmt(X, 0), `Predicted` = fmt(P, 0),
                           Coverage = pct(coverage_95))], caption = "By year"))
  tt <- ev$totals
  print(kbl(data.table(
    Quantity = c("Observed heat deaths, all episodes", "Predicted, published curves",
                 sprintf("Displacement: observed in the %d days after", cfg$post_days),
                 sprintf("Predicted in the %d days after", cfg$post_days)),
    Deaths = c(fmt(tt$observed_excess, 0),
               sprintf("%s (%s to %s)", fmt(tt$predicted, 0), fmt(tt$predicted_lo, 0), fmt(tt$predicted_hi, 0)),
               fmt(tt$post_observed, 0), fmt(tt$post_predicted, 0))),
    caption = "Totals over all heatwave episodes"))
}
```

```{r test-b-calibration, fig.height=5.5}
if (has_ep) {
  w <- ev$windows[type %in% c("episode", "summer")]
  w[, type := factor(type, c("episode", "summer"), c("Heatwave episodes", "Whole summers"))]
  cal <- ev$calibration
  cal[, type := factor(windows, c("Heatwave episodes", "Whole summers (June-September)"),
                       c("Heatwave episodes", "Whole summers"))]
  print(ggplot(w, aes(P, X)) + geom_hline(yintercept = 0, colour = "grey80") +
    geom_abline(colour = "grey40", linetype = 2) +
    geom_abline(data = cal, aes(slope = slope, intercept = 0), colour = "firebrick") +
    geom_errorbarh(aes(xmin = P_lo, xmax = P_hi), colour = "grey75", height = 0) +
    geom_point(aes(colour = factor(year)), alpha = .7) + facet_wrap(~ type, scales = "free") +
    labs(x = "Predicted heat deaths (published curves)", y = "Observed heat deaths (O - B)",
         colour = "Year", title = "Observed vs predicted heat deaths",
         subtitle = "Dashed = perfect calibration; red = fitted slope"))
}
```

```{r test-b-synthesis, results='asis'}
if (has_ep) {
  pt <- tar_read(pooled_test); bt <- tar_read(blup_test)
  rr_ratio <- pt[group == "All cities", ratio]
  ex <- if (!is.null(bt)) (median(bt$results$rr99_blup) - 1) / (median(bt$results$rr99_pub) - 1) else NA
  sl <- ev$calibration[windows == "Heatwave episodes", slope]
  cat(sprintf(paste("\n\n**The two tests agree.** Test A finds refitted heat risks at the 99th",
    "percentile %s times the published ones; on the scale that matters for attributable",
    "deaths (RR - 1) that is a factor of %s. Test B finds a calibration slope of %s.",
    "Two methods that share only the published curves - one fitting a new exposure-response",
    "function, the other counting deaths against a baseline - agree on the answer.\n\n"),
    fmt(rr_ratio, 3), fmt(ex, 2), fmt(sl, 2)))
}
```

The relationship is strong (r = `r fmt(ev$calibration[windows == "Heatwave episodes", r])`
across episodes): the curves rank
episodes correctly, and overstate each of them by roughly the same factor
(`r fmt(ev$calibration[windows == "Heatwave episodes", predicted / observed])`
overall). The effect is visible in the city panel above, where the predicted
daily peak sits well above the observed one while its timing is right. By year,
the largest observed excess is in
`r ev$by_year[type == "episode"][which.max(X), year]` and the smallest ratio of
observed to predicted in
`r ev$by_year[type == "episode"][which.min(X / P), year]`.

```{r test-b-list}
if (has_ep) {
  el <- merge(ev$windows[type == "episode"], tar_read(cities)[, .(URAU_CODE, name)],
              by = "URAU_CODE")[order(-O)]
  kbl(el[1:min(25, .N), .(City = name, Start = start, End = end, O = fmt(O, 0),
                          `Expected without heat` = fmt(B, 0), `Observed heat deaths` = fmt(X, 0),
                          Predicted = sprintf("%s (%s to %s)", fmt(P, 0), fmt(P_lo, 0), fmt(P_hi, 0)),
                          `In 95% PI` = in_pi)],
      caption = "Largest episodes")
}
```

```{r test-b-failures}
if (nrow(cf$failures)) kbl(merge(cf$failures, tar_read(cities)[, .(URAU_CODE, name)], by = "URAU_CODE")[
  , .(City = name, Age = agegroup, Reason = reason)], caption = "Baseline models that needed a fallback or failed")
```

```{r no-episodes, results='asis'}
if (!has_ep) cat("No window could be evaluated.")
```

# Curves re-estimated on Italian data

The published curves are re-estimated on Italian mortality of an earlier
window, with Masselot's own temperatures and his own model, then pooled and
shrunk exactly as in his second stage. Applied to the test period they are
out of sample, since no data after the window enters them. The comparison
below keeps the baseline and every other choice fixed, so it isolates the
effect of the curve itself.

```{r curve-sets}
cs <- tar_read(curve_sets)
kbl(cs$calibration[, .(Windows = fifelse(windows == "episode", "Episodes", "Summers"),
                       Curves = curves, N = n, Observed = fmt(observed, 0),
                       Predicted = fmt(predicted, 0), `Observed / predicted` = fmt(ratio),
                       Slope = sprintf("%s (%s-%s)", fmt(slope), fmt(lo), fmt(hi)),
                       r = fmt(r))],
    caption = "Calibration of published and re-estimated curves (slope 1 = calibrated)")
```

```{r curve-set-numbers}
cse <- cs$calibration[windows == "episode"]
pubr <- cse[curves == "Published (Masselot et al.)"]
recr <- cse[curves != "Published (Masselot et al.)"][order(-ratio)]
```

Re-estimating the level closes most of the gap: the ratio of observed to
predicted goes from `r fmt(pubr$ratio)` with the published curves to
`r paste(sprintf("%s (%s)", fmt(recr$ratio), recr$curves), collapse = " and ")`,
and the slope from `r fmt(pubr$slope)` to
`r paste(fmt(recr$slope), collapse = " and ")`. Which window fits best is
itself informative: a later window fitting better says risk kept falling after
2019, while a longer window fitting better says the shorter one is too noisy.

The slope stays below 1 even where the totals match, so the largest episodes
are still over-predicted relative to the small ones — worth remembering for a
warning system, where the largest episodes are the ones that matter.

```{r example-plot, fig.height=7}
ex <- tar_read(example)
if (!is.null(ex)) {
  d <- ex$data[!is.na(deaths)]
  ggplot(d[series != "Observed"], aes(date, deaths, colour = series)) +
    geom_line(data = d[series == "Observed"], colour = "grey55", alpha = .6, linewidth = .4) +
    geom_line(linewidth = .8) +
    facet_wrap(~ year, ncol = 1, scales = "free_x") +
    scale_colour_manual(values = setNames(
      c("steelblue", "firebrick", c("#1B7837", "#762A83", "#B35806"))[
        seq_len(uniqueN(d[series != "Observed", series]))],
      levels(droplevels(d[series != "Observed", series])))) +
    labs(x = NULL, y = "Deaths per day (20+)", colour = NULL,
         title = sprintf("%s: observed deaths, baseline and predictions", ex$city),
         subtitle = "Grey: observed. Summers of the test period") +
    theme(legend.position = "bottom")
}
```

## Sensitivity: the FluMOMO baseline

Test B measures observed heat deaths against our own baseline of deaths
expected without heat. To check that the result does not depend on it, the
same comparison is repeated against the baseline of the official EuroMOMO
FluMOMO code, run on the comuni of the validated cities of
`r tar_read(config)$flumomo$region` by a different method, at weekly
resolution and for all ages. The expected series there is `EB + EdIA`, the
FluMOMO baseline plus the deaths it attributes to influenza, so influenza
stays out of the residual and what remains above it in the warm season is the
heat signal. Only the warm-season weeks for which the curves produce a
prediction are used, that is those of the test period: FluMOMO is fitted on a
longer history, and earlier weeks would enter with a prediction of zero.

```{r flumomo-sens}
fs2 <- tryCatch(tar_read(flumomo_sensitivity), error = function(e) NULL)
if (is.null(fs2)) {
  cat("Not available: the FluMOMO targets have not been run",
      "(the official code must be unzipped in the folder set in config.R).")
} else {
  kbl(fs2$calibration[, .(Curves = curves, `Summer weeks` = weeks,
                          `Observed heat deaths` = fmt(observed, 0),
                          `Predicted` = fmt(predicted, 0),
                          `Observed / predicted` = fmt(ratio),
                          Slope = sprintf("%s (%s-%s)", fmt(slope), fmt(lo), fmt(hi)),
                          `FluMOMO extreme temperature` = fmt(flumomo_ET, 0))],
      caption = sprintf(paste("Test B against the FluMOMO baseline, %d cities, %s to %s.",
                              "The last column is what FluMOMO itself attributes to",
                              "extreme temperature over the same weeks."),
                        fs2$n_cities, fs2$period[1], fs2$period[2]))
}
```

```{r flumomo-sens-plot, fig.height=4.5}
if (!is.null(fs2)) {
  w <- fs2$weekly[month %in% cfg$warm_months]
  print(ggplot(w, aes(P, X, colour = curves)) +
    geom_hline(yintercept = 0, colour = "grey80") +
    geom_abline(colour = "grey40", linetype = 2) +
    geom_point(alpha = .7) + geom_smooth(method = "lm", formula = y ~ 0 + x, se = FALSE) +
    labs(x = "Predicted heat deaths in the week", y = "Observed above the FluMOMO baseline",
         colour = NULL, title = "Weekly heat deaths against the FluMOMO baseline") +
    theme(legend.position = "bottom"))
}
```

`r if (!is.null(fs2)) "Two baselines built on different principles, at different resolutions, should not give identical numbers; what matters is whether they agree on the direction and the rough size of the gap." else ""`

# Interpretation

```{r interpretation, results='asis'}
pt <- tar_read(pooled_test); cp <- tar_read(curves_periods)
ep <- ev$calibration[windows == "Heatwave episodes"]
rec <- cs$calibration[windows == "episode" & curves != "Published (Masselot et al.)"]
best <- rec[which.min(abs(ratio - 1))]
cat(sprintf(paste0(
"**The published curves are too steep for Italy today.** Our reconstruction ",
"reproduces the paper's own numbers exactly, the cities are aligned to the ",
"population the curves describe, and the temperatures are Masselot's variable. ",
"On that footing, refitting his first-stage model on Italian data gives heat ",
"risks at the 99th percentile %s times the published ones (95%% CI %s-%s), and ",
"counting deaths against a baseline of expected deaths gives a calibration ",
"slope of %s (%s-%s). Over the test period the published curves predict %s heat ",
"deaths against %s observed. The two tests share only the published curves, and ",
"they agree.\n\n"),
fmt(pt[group == "All cities", ratio], 3), fmt(pt[group == "All cities", lo], 3),
fmt(pt[group == "All cities", hi], 3), fmt(ep$slope), fmt(ep$lo), fmt(ep$hi),
fmt(ep$predicted, 0), fmt(ep$observed, 0)))

cat(sprintf(paste0(
"**The gap is not a shape problem and not a place problem.** The refitted ",
"curves track the published ones along the cold arm and separate from them ",
"only above the minimum mortality temperature; heterogeneity between cities is ",
"small; and cities whose curves came from their own mortality behave like ",
"cities where the curve was extrapolated. What is out of date is the level of ",
"heat risk.\n\n")))

cat(sprintf(paste0(
"**It is a time problem.** Window by window the ratio falls: %s, a change of ",
"%s per decade (95%% CI %s-%s, p = %s), with the two earlier windows using ",
"Masselot's own temperatures. The Italian mortality behind the published ",
"curves ends in 2010 and includes the 2003 heatwave, so these curves describe ",
"a population that was more vulnerable to heat than today's. Warning systems, ",
"air conditioning and changes in the care of frail elderly people are the usual ",
"explanations, and the pattern here is consistent with them.\n\n"),
paste(sprintf("%s in %s", fmt(cp$pooled$ratio, 3), cp$pooled$window), collapse = ", "),
fmt(cp$trend$change_per_decade, 3), fmt(cp$trend$lo, 3), fmt(cp$trend$hi, 3),
fmt(cp$trend$p, 3)))

cat(sprintf(paste0(
"**Re-estimating the level repairs most of it, out of sample.** Curves fitted ",
"on Italian data that end before the test period bring the observed-to-",
"predicted ratio from %s to %s (\"%s\"), with a slope of %s. The framework is ",
"sound: the model form, the exposure, the baseline and the attribution all ",
"work; the published numbers simply carry the risk level of the early 2000s. ",
"For operational use the recalibrated curves are the ones to adopt, refreshed ",
"as new years accumulate.\n\n"),
fmt(ep$observed / ep$predicted), fmt(best$ratio), best$curves, fmt(best$slope)))

cat(sprintf(paste0(
"**Displacement is part of the story.** In the %d days after an episode the ",
"published curves predict %s further deaths while %s are observed above the ",
"baseline, so a large part of what the curves count is death brought forward ",
"by days rather than added. A cumulative 21-day curve carries this inside it, ",
"which is one reason a curve fitted on longer series transfers better than a ",
"short one.\n\n"), cfg$post_days, fmt(ev$totals$post_predicted, 0),
fmt(ev$totals$post_observed, 0)))

if (!is.null(fs2)) {
  fpub <- fs2$calibration[curves == "Published (Masselot et al.)"]
  closer <- abs(log(pmax(fpub$flumomo_ET, 1) / fpub$predicted)) <
    abs(log(pmax(fpub$observed, 1) / fpub$predicted))
  cat(sprintf(paste0(
  "**A second baseline.** Repeating the comparison against the baseline of the ",
  "official FluMOMO code, on %d %s at weekly resolution, gives a slope of %s ",
  "for the published curves (observed over predicted %s), against %s in the ",
  "main test. FluMOMO itself attributes %s deaths to extreme temperature over ",
  "those weeks, against %s predicted by the published curves, so %s\n\n"),
  fs2$n_cities, ifelse(fs2$n_cities == 1, "city,", "cities,"),
  fmt(fpub$slope), fmt(fpub$ratio), fmt(ep$slope), fmt(fpub$flumomo_ET, 0),
  fmt(fpub$predicted, 0),
  if (closer) paste("the two model-based estimates sit closer to each other than",
                    "either does to the observed excess.")
  else paste("the two model-based estimates differ from each other by as much as",
             "the published curves differ from the observed excess.")))
}

cat(sprintf(paste0(
"**What is not settled.** Part of the recent decline may not be adaptation: ",
"the 2020-2022 excess mortality removed frail people from the population, ",
"which lowers heat susceptibility in 2023-2025 and should partly revert. The ",
"published curves are also meta-regression predictions, so some of the gap ",
"could be model bias present from the start rather than ageing of the ",
"estimates; the earliest window, already at %s, is consistent with a small ",
"constant component of that kind.\n\n"), fmt(cp$pooled$ratio[1], 2)))

cat(paste0(
"**For a forecasting system.** Use recalibrated Italian curves rather than the ",
"published ones; debias the temperature forecasts before converting them into ",
"deaths, since the warm bias on hot days inflates predicted deaths at short ",
"leads; and treat the output as a ranking of episode severity, which is well ",
"calibrated, more than as an exact death count, which is not.\n\n"))
```

# Limitations

- **Place of residence.** Istat counts deaths by comune of residence, not of
  occurrence, as do the mortality series used by Masselot et al.
- **Same-day attribution.** The published curves are overall cumulative
  effects over lags 0-`r cfg$lag_max`. Test B attributes each day's
  cumulative risk to that day; episode windows include
  `r cfg$tail_days` extra days to absorb most of the lagged deaths, and the
  post-episode excess is reported to show displacement.
- **Exposure.** Our series is taken at the city point, Masselot's is a
  city-polygon average; the difference is measured and
  `r if (isTRUE(cfg$bias_correct)) "removed as a monthly mean" else "not corrected"`.
  Forecasts come from a coarser model grid than ERA5-Land.
- **Uncertainty.** `r if (isTRUE(cfg$use_coef_simu)) "Published coefficient draws (coef_simu.csv) keep the correlation between age groups and cities." else "Coefficient draws come from vcov.csv, which ignores the correlation between age groups and between cities; national intervals are therefore too narrow. Setting use_coef_simu = TRUE fixes this."`
- **Power.** With about three and a half years, small cities cannot support
  the first-stage model and are left to the pooled analyses and test B.
- **Post-pandemic period.** 2023 still carried residual effects of
  COVID-19 and of the mortality deficit that followed it; the spline of time
  absorbs slow changes but not short epidemics during episodes.
- **Recalibration.** The recalibrated curves are fitted on all deaths aged 20+,
  not by age group, and keep the published cold side inside the baseline;
  both are refinements for a later version.

# Configuration

```{r config}
kbl(data.table(Parameter = names(cfg),
               Value = vapply(cfg, function(v) paste(format(v), collapse = ", "), "")))
```
